---
title: "Assignment 3: Model comparison"
author: "Marton Kovacs"
output: html_document
editor_options: 
  chunk_output_type: console
---

In this lab assignment you are going to work with (simulated) data related to perioperative pain and its psychological and hormonal predictors. In the assignment you will assess the added benefit of including some psychological and hormonal predictors to the already established demographic predictors of pain.

In this assignment you will set up a hierarchical regression model to predict postoperative pain after wisdom tooth surgery. 

# Research problem

The amount of pain experienced around and after surgeries are highly variable between and within individuals. In order to improve surgical pain management regimens we need to understand what influences pain around surgical procedures and predict the amount of pain an individual will experience.

Your first study in this area is related to assessing the influence of trait and state psychological measures on pain, and to see whether taking into account these variables can improve our understanding of postoperative pain.

# Procedures and measures

Use the data file called ‘assignment_3_dataset’, from the 'data/' folder.

You have collected data from 160 adults who were scheduled to undergo surgical extraction of the third mandibular molar (wisdom tooth surgery). Patients filled out a form in the waiting room before their surgery. The form contained questions about their sex, age, and weight, and psychological questionnaires assessing anxiety, pain catastrophizing, and mindfulness (see descriptions below). You also got blood samples and saliva samples from participants in the waiting room 5 minutes before their operations to determine the serum (a component of the blood) and salivary cortisol levels of participants. Participants were contacted 5 hours after the surgery to see how much pain they were experiencing. The __level of pain__ at that moment was recorded using a numerical rating scale using a __scale of 0 to 10__, where 0 means “no pain” and 10 means “worst pain I can imagine”. 

__The State Trait Anxiety Inventory:__ T measures trait anxiety on a scale of 20 to 80, higher scores mean higher anxiety. Anxiety has been found in many studies to positively correlate with the level of pain experienced. This is __variable STAI_trait__ in the dataset.

__The Pain Catastrophizing Scale__ measures the extent of pain catastrophizing, which is characterized by a tendency to magnify the threat value of a pain stimulus and to feel helpless in the presence of pain, as well as by a relative inability to prevent or inhibit pain-related thoughts in anticipation of, during, or following a painful event. The total score on this scale ranges from 0 to 52, higher scores mean higher catastrophizing. Pain catastrophizing is one of the well-established predictors of clinical pain. This is __variable pain_cat__ in the dataset.

__The Mindful Attention Awareness Scale (MAAS)__ measures dispositional mindfulness, which may be described as a tendency to turn attention to present-moment experiences in an open, non-judgmental way. The MAAS total score ranges from 1 to 6 (an average of the item scores), with higher scores representing higher dispositional mindfulness. Trait mindfulness has been theorized to serve as a protective factor against pain, as the individual would be more objective about their pain experience and tend to associate less discomfort, despair, and hopelessness to the pain-related sensations. This is __variable mindfulness__ in the dataset.

__Cortisol__ is a stress hormone associated with acute and chronic stress. Cortisol levels are thought to be positively associated with pain experience. Cortisol can be __measured from both blood and the saliva__, although, serum cortisol is often regarded in medical research as more reliably related to stress (serum is a component of the blood plasma). These are __variables cortisol_serum__, and __cortisol_saliva__ in the dataset.

# Research question

Previous studies and meta-analyses showed that age and sex are often predictors of pain (age is negatively associated with pain, while sex is a predictor more dependent on the type of the procedure). You would like to determine the extent to which taking into account psychological and hormonal variables aside from the already used demographic variables would improve our understanding of postoperative pain.

To answer this research question you will __need to compare two models__ (with a hierarchical regression). The __simpler model__ should contain __age and sex as predictors of pain__, while the __more complex model__ should contain the __predictors: age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures__. Notice that the predictors used in the simpler model are a subset of the predictors used in more complex model. __You will have to do model comparison to assess whether substantial new information was gained about pain in the more complex model compared to the simpler model.__  

# What to report

As usual, before you can interpret your model, you will need to run data and model diagnostics. First, check the variables included in the more complex model (age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures as predictors, and pain as an outcome) for __coding errors__, and the model itself for __influential outliers__ (for example using Cook’s distance). Furthermore, check the final model to see if the __assumptions of linear regression hold true__, that is, __normality__ (of the residuals), __linearity__ (of the relationship), __homogeneity of variance__ (also called homoscedasticity) and that there is no excess __multicollinearity__ (“uncorrelated predictors” in Navarro’s words). If you find anything amiss during these checks, make the appropriate decision or correction and report your findings and actions in your report. 

__Note:__ If you do any changes, such as exclude cases, or exclude predictors from the model, you will have to re-run the above checks for your final data and model.

Report the results of the simpler model and the more complex model. For both models you should report the model test statistics (adj.R2, F, df, and p value). Also, report the statistics describing the coefficients of the predictors in a table format (unstandardized regression coefficients and 95% confidence intervals, standardized regression coefficients (B and Beta values), and p values).

Write up the regression equation of the more complex model in the form of 𝑌 = 𝑏0 + 𝑏1 ∗ X1 + 𝑏2 ∗ X2 +…+ bn * Xn, in which you use the actual regression coefficients of your models. (b0 stands for the intercept and b1, b2 … bn stand for the model coefficients for each of the predictors, and X1, X2, … Xn denote the predictors).

Compare the two models in terms of how much variance they explain of pain’s variability in the sample. Report Akaike information criterion (AIC) for both models and the F test statistic and p value of the likelihood ratio test comparing the two models.

# What to discuss

In your discussion of the findings, briefly interpret the results of the above analyses, and indicate whether you think that anything was gained by including the psychological and hormone measures in the model.

# Solution

## Read the data

Read the dataset used in this assignment. Pay attention to the extension of the datafile.

```{r}
require(readxl)

pain_data <- read_excel("data/assignment_3_dataset.xlsx")

```

## Data and model diagnostics 
### Data diagnostics
#### Descriptives of the variables

Run an exploratory data analysis (EDA) to investigate the dataset.

```{r}

require(tidyverse)

#we check the data, by plotting histograms of each numeric column and summarising the sex column
pain_data %>%
  group_by(sex) %>%
  summarise(count = n())

pain_data %>%
  select(where(is.numeric)) %>%
  gather() %>%
  ggplot( aes(value)) +
    geom_histogram(bins=8) +
    facet_wrap(~key, scales = 'free_x')

pain_data %>%
  map(summary)

```

#### Correct coding errors

If you find values in the dataset during the EDA, that are not correct based on the provided descriptions of the variables of the dataset please correct them here.

```{r}

#we already see, that the sex variable has 1 woman type, while the others are female/male coded. We recode this value. Then we also recode into numerical varibale: 0 = female, 1 = male
#from the summary tables, we can see that the maximum of the Mindfulness Scale is 6.22, which should not be the case, as it ranges from 1 to 6. We exclude this value
#in the histogram plots, the pain scale does not look normal, there is an extreme outlier of value 50, which should be a coding error. As we are not entirely sure whether it was meant to be a 5,so  we exclude the participant

pain_data_cleaned = pain_data %>%
  mutate(sex = sub("woman", "female", sex)) %>%
  mutate(sex = ifelse(sex=="female", 0, 1)) %>%
  mutate(sex = factor(sex, levels=c(0, 1), labels = c("female","male"))) %>%
  filter(pain <= 10) %>%
  filter(mindfulness <= 6)

#we now check the cleaned dataset
pain_data_cleaned %>%
  group_by(sex) %>%
  summarise(count = n())

pain_data_cleaned %>%
  select(where(is.numeric)) %>%
  gather() %>%
  ggplot( aes(value)) +
    geom_histogram(bins=8) +
    facet_wrap(~key, scales = 'free_x')

pain_data_cleaned %>%
  map(summary)



```

### Model diagnostics
#### Build the more complex model

In order to test the more complex model for outliers and to test the assumptions first build the model.

```{r}
#age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures

complex_model = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum + cortisol_saliva, pain_data_cleaned)
summary(complex_model)

```

#### Checking for influential outliers

Check for outlier values in the model.

```{r}

# Plot the Cook's Distance using the traditional 4/n criterion - based on: http://r-statistics.co/Outlier-Treatment-With-R.html
cook_distance_remove <- function(some_model, some_data) {
  cookdistance = cooks.distance(some_model)

  sample_size <- nrow(some_data)
  plot(cookdistance, pch="*", cex=2, main="Cook Distance influential outliers")  # plot cook's distance
  abline(h = 4/sample_size, col="red")  # add cutoff line
  text(x=1:length(cookdistance)+1, y=cookdistance, labels=ifelse(cookdistance>4/sample_size, names(cookdistance),""), col="red")  # add labels
  
  influential <- as.numeric(names(cookdistance)[(cookdistance > 4*mean(cookdistance, na.rm=T))])
  head(some_data[influential, ])
  
  some_data_outlierrm = some_data %>%
    slice(-influential)
  return(some_data_outlierrm)
}

#we run the cook's distance outlier removal several times, until we no longer find influential values.
pain_data_cleaned_outlierrm = cook_distance_remove(complex_model, pain_data_cleaned)
complex_model2 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum + cortisol_saliva, pain_data_cleaned_outlierrm)

pain_data_cleaned_outlierrm2 = cook_distance_remove(complex_model2, pain_data_cleaned_outlierrm)
complex_model3 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum + cortisol_saliva, pain_data_cleaned_outlierrm2)

pain_data_cleaned_outlierrm3 = cook_distance_remove(complex_model3, pain_data_cleaned_outlierrm2)
complex_model4 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum + cortisol_saliva, pain_data_cleaned_outlierrm3)

pain_data_cleaned_outlierrm4 = cook_distance_remove(complex_model4, pain_data_cleaned_outlierrm3)
complex_model5 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum + cortisol_saliva, pain_data_cleaned_outlierrm4)

pain_data_cleaned_outlierrm5 = cook_distance_remove(complex_model5, pain_data_cleaned_outlierrm4)
complex_model6 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum + cortisol_saliva, pain_data_cleaned_outlierrm5)


```

#### Checking assumptions

Check the normality assumption.

```{r}
#Q-Q plot
plot(complex_model6, 2)
#Histogram
residuals_complex_model6 = enframe(residuals(complex_model6))
residuals_complex_model6 %>% ggplot() + aes(x=value) + geom_histogram()

#Normality is slightly violated, which could be due to the few amount of observations
#We choose to use a stricter significance cireria of <0.01 (according to https://github.com/kekecsz/PSYP14-Advanced-Scientific-Methods/blob/main/Exercise_11_Model_diagnostics/Exercise_11_Model_diagnostics.pdf)

```

Check the linearity assumption.

```{r}
require(car)

plott = plot(complex_model6, 1)
complex_model6 %>% residualPlots()

#As none of the tests are significant, even though we see a bit of a curvature, we assume linaerity holds.
  
```

Check the homoscedasticty assumption (homogeneity of variance).

```{r}
require(lmtest)

#We plot the stansardized residuals which show a curious pattern, which can be attributed to the ordinality of the dependent variable
plot(complex_model6, 3)

#We test significant heteroscdacity with NCV and BT tests based on https://github.com/kekecsz/PSYP14-Advanced-Scientific-Methods/blob/main/Exercise_11_Model_diagnostics/Exercise_11_Model_diagnostics.pdf
complex_model6 %>% ncvTest()
complex_model6 %>% bptest()

#We see no significant heteroscedacity, so homoscedacity of the data can be assumed

```

Check the multicollinearity assumption.

(VIF above 5), or a VIF threshold of 3 is recommended in this paper: http://onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2009.00001.x/full

Some info about VIF: 
https://statisticalhorizons.com/multicollinearity
http://blog.minitab.com/blog/understanding-statistics/handling-multicollinearity-in-regression-analysis

```{r}
require(psych)

#Based on the suggested readings, we treat variables above VIF of 3.
#There are two values like this: cortisol_serum and cortisol_saliva
#These two measures are different methods to check the cortisol levels, so they essentially give the same information.
car::vif(complex_model6)

pain_data_cleaned_outlierrm5 %>% select(cortisol_serum, cortisol_saliva) %>%
  pairs.panels(col="red",lm=T)

#We choose to drop cortisol_saliva and use only cortisol_serum as a cortisol metric in our model, because saliva is a more indirect measure of cortisol than blood serum.

```

### Making decision based on model diagnostics

If based on the assumption tests you decide to drop a predictor variable you should do that here. Create your updated model.

```{r}
#based on the VIF analysis we remove the cortisol_saliva variable
complex_model_nosaliva = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum, pain_data_cleaned)
summary(complex_model_nosaliva)

```

#### Checking outliers of the updated model

```{r}
#we run the cook's distance outlier removal several times, until we no longer find influential values.
pain_data_cleaned_nosaliva_outlierrm = cook_distance_remove(complex_model_nosaliva, pain_data_cleaned)
complex_model_nosaliva2 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum, pain_data_cleaned_nosaliva_outlierrm)

pain_data_cleaned_nosaliva_outlierrm2 = cook_distance_remove(complex_model_nosaliva2, pain_data_cleaned_nosaliva_outlierrm)
complex_model_nosaliva3 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum, pain_data_cleaned_nosaliva_outlierrm2)

pain_data_cleaned_nosaliva_outlierrm3 = cook_distance_remove(complex_model_nosaliva3, pain_data_cleaned_nosaliva_outlierrm2)
complex_model_nosaliva4 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum, pain_data_cleaned_nosaliva_outlierrm3)

pain_data_cleaned_nosaliva_outlierrm4 = cook_distance_remove(complex_model_nosaliva4, pain_data_cleaned_nosaliva_outlierrm3)
complex_model_nosaliva5 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum, pain_data_cleaned_nosaliva_outlierrm4)

pain_data_cleaned_nosaliva_outlierrm5 = cook_distance_remove(complex_model_nosaliva5, pain_data_cleaned_nosaliva_outlierrm4)
complex_model_nosaliva6 = lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum, pain_data_cleaned_nosaliva_outlierrm5)
```

#### Checking assumptions of the updated model

Normality assumption

```{r}
#Q-Q plot
plot(complex_model_nosaliva6, 2)
#Histogram
residuals_complex_model_nosaliva6 = enframe(residuals(complex_model_nosaliva6))
residuals_complex_model_nosaliva6 %>% ggplot() + aes(x=value) + geom_histogram()

#Normality is slightly violated, which could be due to the few amount of observations
#We choose to use a stricter significance cireria of <0.01 (according to https://github.com/kekecsz/PSYP14-Advanced-Scientific-Methods/blob/main/Exercise_11_Model_diagnostics/Exercise_11_Model_diagnostics.pdf)
```

Linearity assumption

```{r}
plott = plot(complex_model_nosaliva6, 1)
complex_model_nosaliva6 %>% residualPlots()

#As none of the tests are significant, even though we see a bit of a curvature, we assume linaerity holds.
```

Homoscedasticty assumption (homogeneity of variance)

```{r}
#We plot the stansardized residuals which show a curious pattern, which can be attributed to the ordinality of the dependent variable
plot(complex_model_nosaliva6, 3)

#We test significant heteroscdacity with NCV and BT tests based on https://github.com/kekecsz/PSYP14-Advanced-Scientific-Methods/blob/main/Exercise_11_Model_diagnostics/Exercise_11_Model_diagnostics.pdf
complex_model_nosaliva6 %>% ncvTest()
complex_model_nosaliva6 %>% bptest()

#We see no significant heteroscedacity, so homoscedacity of the data can be assumed
```

Multicollinearity assumption

```{r}
#Based on the suggested readings, we treat variables above VIF of 3.
#In this new model, we see no multicolleniarity
car::vif(complex_model_nosaliva6)
```

## Model comparison

Create the simple model and get the results of the model that needs to be reported based on the What to report section.

The model's test statistics are:
Adjusted R-squared = 0.06712
F-statistic = 5.713
df = 2 and 129
p value =  0.004195

Based on the coefficients, we see that age (b = -0,072; p = 0.001) is a significant predictor of subjective feeling of pain around and after surgical procedures.

```{r}
require(lsr)

#We create the simple model with only age and sex as predictors
simple_model = lm(pain ~ age + sex, pain_data_cleaned_nosaliva_outlierrm5)
summary(simple_model)

#Getting the relevant data of predictors in a table 
sum_simp = summary(simple_model)[["coefficients"]] #unstandardized coefficients, t and p
conf_simp = confint(simple_model) #95% confidence intervals
betas_simp = standardCoefs(simple_model) #standardized coefficients (B and beta)

simple_model_coef_table = cbind(sum_simp[-1,], conf_simp[-1,], betas_simp)
simple_model_coef_table

```

Create the more complex model based on the results of the model diagnostics. Also, get the results that needs to be reported based on the What to report section.

The model's test statistics are:
Adjusted R-squared = 0.4766
F-statistic = 20.88
df = 6 and 125
p value < 2.2e-16

The model's equation:
Y = 0.7 - 0.026 * age + 0.24 * sex_male + 0.216 * STAI_trait + 0.8616 * pain_catastrophizing - 0.155 * mindfulness + 0.428 * cortisol_serum

Based on the coefficients, we see that pain catastrophizing (b = 0.862; p = 0.0007) and cortisol measured from serum (b = 0.428; p = 0.0001) are significant predictors of subjective feeling of pain around and after surgical procedures.

```{r}

#We use the final version of complex model without influential outliers and checked for all assumptions
summary(complex_model_nosaliva6)

#Getting the relevant data of predictors in a table 
sum_comp = summary(complex_model_nosaliva6)[["coefficients"]] #unstandardized coefficients, t and p
conf_comp = confint(complex_model_nosaliva6) #95% confidence intervals
betas_comp = standardCoefs(complex_model_nosaliva6) #standardized coefficients (B and beta)

complex_model_coef_table = cbind(sum_comp[-1,], conf_comp[-1,], betas_comp)
complex_model_coef_table

```

Compare the two models.

Compare the two models in terms of how much variance they explain of pain’s variability in the sample. Report Akaike information criterion (AIC) for both models and the F test statistic and p value of the likelihood ratio test comparing the two models.

```{r}
require(lmtest)
require(stats)

#Comparing the two models with likelihood ratio test
anova(simple_model, complex_model_nosaliva6)
# F-test: 26.226; p = 8.517e-16
lrtest(simple_model, complex_model_nosaliva6)
# ChiSq = 80.434; p < 2.2e-16

#Reporting the AIC of the models
AIC(simple_model) # Simple_model AIC = 432.852
AIC(complex_model_nosaliva6) # Complex model AIC = 360.418

AIC(simple_model, complex_model_nosaliva6)

```

Comparing the two models we see that the complex model is significantly better at explaining the variance (F = 26.226; p = 8.517e-16), than the simple model. 
The complex model is also a better fit to the data based on the AIC-measueres (complex AIC = 360.418; simple AIC = 432.852) 
