---
title: "Assignment 3: Model comparison"
author: "Marton Kovacs"
output: html_document
editor_options: 
  chunk_output_type: console
---

In this lab assignment you are going to work with (simulated) data related to perioperative pain and its psychological and hormonal predictors. In the assignment you will assess the added benefit of including some psychological and hormonal predictors to the already established demographic predictors of pain.

In this assignment you will set up a hierarchical regression model to predict postoperative pain after wisdom tooth surgery. 

# Research problem

The amount of pain experienced around and after surgeries are highly variable between and within individuals. In order to improve surgical pain management regimens we need to understand what influences pain around surgical procedures and predict the amount of pain an individual will experience.

Your first study in this area is related to assessing the influence of trait and state psychological measures on pain, and to see whether taking into account these variables can improve our understanding of postoperative pain.

# Procedures and measures

Use the data file called ‘assignment_3_dataset’, from the 'data/' folder.

You have collected data from 160 adults who were scheduled to undergo surgical extraction of the third mandibular molar (wisdom tooth surgery). Patients filled out a form in the waiting room before their surgery. The form contained questions about their sex, age, and weight, and psychological questionnaires assessing anxiety, pain catastrophizing, and mindfulness (see descriptions below). You also got blood samples and saliva samples from participants in the waiting room 5 minutes before their operations to determine the serum (a component of the blood) and salivary cortisol levels of participants. Participants were contacted 5 hours after the surgery to see how much pain they were experiencing. The __level of pain__ at that moment was recorded using a numerical rating scale using a __scale of 0 to 10__, where 0 means “no pain” and 10 means “worst pain I can imagine”. 

__The State Trait Anxiety Inventory:__ T measures trait anxiety on a scale of 20 to 80, higher scores mean higher anxiety. Anxiety has been found in many studies to positively correlate with the level of pain experienced. This is __variable STAI_trait__ in the dataset.

__The Pain Catastrophizing Scale__ measures the extent of pain catastrophizing, which is characterized by a tendency to magnify the threat value of a pain stimulus and to feel helpless in the presence of pain, as well as by a relative inability to prevent or inhibit pain-related thoughts in anticipation of, during, or following a painful event. The total score on this scale ranges from 0 to 52, higher scores mean higher catastrophizing. Pain catastrophizing is one of the well-established predictors of clinical pain. This is __variable pain_cat__ in the dataset.

__The Mindful Attention Awareness Scale (MAAS)__ measures dispositional mindfulness, which may be described as a tendency to turn attention to present-moment experiences in an open, non-judgmental way. The MAAS total score ranges from 1 to 6 (an average of the item scores), with higher scores representing higher dispositional mindfulness. Trait mindfulness has been theorized to serve as a protective factor against pain, as the individual would be more objective about their pain experience and tend to associate less discomfort, despair, and hopelessness to the pain-related sensations. This is __variable mindfulness__ in the dataset.

__Cortisol__ is a stress hormone associated with acute and chronic stress. Cortisol levels are thought to be positively associated with pain experience. Cortisol can be __measured from both blood and the saliva__, although, serum cortisol is often regarded in medical research as more reliably related to stress (serum is a component of the blood plasma). These are __variables cortisol_serum__, and __cortisol_saliva__ in the dataset.

# Research question

Previous studies and meta-analyses showed that age and sex are often predictors of pain (age is negatively associated with pain, while sex is a predictor more dependent on the type of the procedure). You would like to determine the extent to which taking into account psychological and hormonal variables aside from the already used demographic variables would improve our understanding of postoperative pain.

To answer this research question you will __need to compare two models__ (with a hierarchical regression). The __simpler model__ should contain __age and sex as predictors of pain__, while the __more complex model__ should contain the __predictors: age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures__. Notice that the predictors used in the simpler model are a subset of the predictors used in more complex model. __You will have to do model comparison to assess whether substantial new information was gained about pain in the more complex model compared to the simpler model.__  

# What to report

As usual, before you can interpret your model, you will need to run data and model diagnostics. First, check the variables included in the more complex model (age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures as predictors, and pain as an outcome) for __coding errors__, and the model itself for __influential outliers__ (for example using Cook’s distance). Furthermore, check the final model to see if the __assumptions of linear regression hold true__, that is, __normality__ (of the residuals), __linearity__ (of the relationship), __homogeneity of variance__ (also called homoscedasticity) and that there is no excess __multicollinearity__ (“uncorrelated predictors” in Navarro’s words). If you find anything amiss during these checks, make the appropriate decision or correction and report your findings and actions in your report. 

__Note:__ If you do any changes, such as exclude cases, or exclude predictors from the model, you will have to re-run the above checks for your final data and model.

Report the results of the simpler model and the more complex model. For both models you should report the model test statistics (adj.R2, F, df, and p value). Also, report the statistics describing the coefficients of the predictors in a table format (unstandardized regression coefficients and 95% confidence intervals, standardized regression coefficients (B and Beta values), and p values).

Write up the regression equation of the more complex model in the form of 𝑌 = 𝑏0 + 𝑏1 ∗ X1 + 𝑏2 ∗ X2 +…+ bn * Xn, in which you use the actual regression coefficients of your models. (b0 stands for the intercept and b1, b2 … bn stand for the model coefficients for each of the predictors, and X1, X2, … Xn denote the predictors).

Compare the two models in terms of how much variance they explain of pain’s variability in the sample. Report Akaike information criterion (AIC) for both models and the F test statistic and p value of the likelihood ratio test comparing the two models.

# What to discuss

In your discussion of the findings, briefly interpret the results of the above analyses, and indicate whether you think that anything was gained by including the psychological and hormone measures in the model.

# Solution

## Read the data

Read the dataset used in this assignment. Pay attention to the extension of the datafile.

```{r}


install.packages("skimr")
install.packages("psych")
install.packages("car")
install.packages("Hmisc")
install.packages("rstatix")
install.packages("olsrr")
install.packages("lmtest")
library(readxl)
library(tidyverse)
library(skimr)
library(psych)
library(car)
library(Hmisc)
library(rstatix)
library(olsrr)
library(lmtest)

pain2 <- read_excel("C:\\Users\\Ma. de Lourdes\\OneDrive - Kormányzati Informatikai Fejlesztési Ügynökség\\UNIVERSIDAD\\PhD\\R Projects\\public_r_data_analysis_2021_spring\\data\\assignment_3_dataset.xlsx",  sheet = 1)

pain_data <- as.data.frame(pain2)

pain_data$sex <- as.factor(pain_data$sex)


```

## Data and model diagnostics 
### Data diagnostics
#### Descriptives of the variables

Run an exploratory data analysis (EDA) to investigate the dataset.

```{r}
# We explore the dataset. Check the type of variables, we skim it, check minimum, maximum and mean values of each variable 

view(pain_data)
str(pain_data)
summary(pain_data)
sum(is.na(pain_data))

skim(pain_data)


# We create a scatterplot to visualize the relationship between age and pain, adding sex as color.

ggplot(data = pain_data) +
  geom_point(mapping = aes(x = age, y = pain, color = sex))
# We visualize relationship between one of the pain measurements and pain, adding sex as color. In both graphs is clear that pain has an observation of 50 which is more thant the range accepted. Also the variable sex has three levels when it should be two

ggplot(pain_data, aes(x = pain_cat, y = pain, color = sex)) +
  geom_point()  +
  geom_smooth(method = "lm")

```

#### Correct coding errors

If you find values in the dataset during the EDA, that are not correct based on the provided descriptions of the variables of the dataset please correct them here.

```{r}

# After exploring the dataset we identified several coding errors. 
# We clean the data of what we interpreted as typing errors and in the case of the variable mindfulness since maximum range is 6, we are not sure whether it was meant to be 6 or another value, so we exclude it.
pain_data_c <- pain_data %>%
  mutate(sex = recode_factor(sex, "woman" = "female")) %>%
  filter(mindfulness <= 6)

#We correct the value in pain variable, as the 0 is most likely to be a error in entry.

pain_data_c$pain[pain_data_c$pain == 50] <- 5

# We explore the cleaned dataset to verify changes.

view(pain_data_c)
summary(pain_data_c)
str(pain_data_c)

# We plot again some of variables to see the differences after correcting coding errors

# We reproduce the same plot as before, we verify that pain has not higher values than 10 
ggplot(data = pain_data_c) +
  geom_point(mapping = aes(x = age, y = pain, color = sex))

# We visualize other variables to check their interactions and possible outliers
ggplot(pain_data_c, aes(x = cortisol_serum, y = pain_cat, fill = sex)) +
  geom_boxplot()  

ggplot(pain_data_c, aes(x = cortisol_saliva, y = pain)) +
  geom_point(color = "aquamarine4") +
  facet_wrap(~ sex)

# We plot minfulness and pain, we verify that the there is no observation higher than 6
ggplot(pain_data_c, aes(x = mindfulness, y = pain)) +
  geom_point(color = "aquamarine4") +
  facet_wrap(~ sex)

```

### Model diagnostics
#### Build the more complex model

In order to test the more complex model for outliers and to test the assumptions first build the model.

```{r}

# Step 1 Block 1 : Nested model

# We build the complex model in which we include the mentioned variables. 

c_model <- lm(pain ~ age + sex + STAI_trait + pain_cat + cortisol_serum + cortisol_saliva + mindfulness, data = pain_data_c)
nrow(model.frame(c_model))
summary(c_model)

```

#### Checking for influential outliers

Check for outlier values in the model.

```{r}
# We use Cook's ditance to estimate the influence of the observations

cooksd <- cooks.distance(c_model)

# We plot the Cook's Distance using the traditional 4/n criterion
sample_size <- nrow(pain_data_c)

plot(cooksd, pch="*", cex=2, main="Influential Obs by Cooks distance")  # plot cook's distance
abline(h = 4/sample_size, col="red")  # add cutoff line
text(x=1:length(cooksd)+1, y=cooksd, labels=ifelse(cooksd>4/sample_size, names(cooksd),""), col="red")  # add labels

# We define outliers based on 4/n criteria; 

influential <- as.numeric(names(cooksd)[(cooksd > (4/sample_size))])

# We visualize the rows that are considered outliers and will be later excluded

view(pain_data_c[influential, ])

# We define the dataset without outliers
pain_data2 <- pain_data_c[-influential, ]
head(pain_data2)

nrow(pain_data2)

# We convert the dataset in data frame for further anaylsis

pain_no_out <- as.data.frame(pain_data2)

view(pain_no_out)


# We build the same complex model, but this time with the dataframe without outliers

c_model_no <- lm(pain ~ age + sex + STAI_trait + pain_cat + cortisol_serum + cortisol_saliva + mindfulness, data = pain_no_out)
nrow(model.frame(c_model_no))
summary(c_model_no)

#Not sure about this plot
outlier_values2 <- boxplot.stats(pain_no_out$pain_cat)$out  # outlier values.
boxplot(pain_no_out$pain_cat, main="Pressure Height", boxwex=0.1)
mtext(paste("Outliers: ", paste(outlier_values2, collapse=", ")), cex=0.6)

```

#### Checking assumptions

Check the normality assumption.

```{r}

#The Shapiro-Wilk statistic associated with these data is W .95, indicating that no significant departures from normality were detected (p 8.47).
shapiro.test(pain_no_out$pain)

# However when plotting normal distribution seems slightly violated

qqnorm(pain_no_out$pain,main="QQ plot of normal data",pch=19)

hist(pain_no_out$pain,probability=T, main="Histogram of normal
data",xlab="Approximately normally distributed data")
lines(density(pain_no_out$pain),col=2)


# Plot for detecting violation of normality assumption
ols_plot_resid_qq(c_model_no)

plot(c_model_no, 2)
# Test for detecting violation of normality assumption.

ols_test_normality(c_model_no)


#ols_test_correlation(c_model)
#ols_plot_resid_hist(c_model)
```

Check the linearity assumption.

```{r}
# Plot to check linearity
ols_plot_resid_fit(c_model_no)

plot(c_model_no, 1)

# Is it linear?.

```

Check the homoscedasticty assumption (homogeneity of variance).

```{r}
ols_plot_resid_fit(c_model_no)

plot(c_model_no, 3)

ols_test_breusch_pagan(c_model_no)
```

Check the multicollinearity assumption.

(VIF above 5), or a VIF threshold of 3 is recommended in this paper: http://onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2009.00001.x/full

Some info about VIF: 
https://statisticalhorizons.com/multicollinearity
http://blog.minitab.com/blog/understanding-statistics/handling-multicollinearity-in-regression-analysis

```{r}

ols_coll_diag(c_model_no)

ols_vif_tol(c_model_no)

vif(c_model_no)

# Cortisol serum and cortisol saliva have high indexes of multicollarity, this means that they are highly correlated predictors, its necessary to remover one for a better model fit. 

# Base on some articles Cortisol saliva is a reliable measure of cortisol levels. 10.1378/chest.11-0448, 10.1177/000456328302000601, https://academic.oup.com/painmedicine/article/9/6/752/1909385#29385949
# We keep cortisol saliva variable in the complex model

```

### Making decision based on model diagnostics

If based on the assumption tests you decide to drop a predictor variable you should do that here. Create your updated model.

```{r}

# We create the updated complex model

c_modelupdate <- lm(pain ~ age + sex + STAI_trait + pain_cat + cortisol_saliva + mindfulness, data = pain_data_c)
nrow(model.frame(c_modelupdate))
summary(c_modelupdate)
```

#### Checking outliers of the updated model

```{r}
# We use Cook's ditance to estimate the influence of the observations

cooksd2 <- cooks.distance(c_modelupdate)

# We plot the Cook's Distance using the traditional 4/n criterion

sample_size2 <- nrow(pain_data_c)

plot(cooksd2, pch="*", cex=2, main="Influential Obs by Cooks distance")  # plot cook's distance
abline(h = 4/sample_size2, col="red")  # add cutoff line
text(x=1:length(cooksd2)+1, y=cooksd2, labels=ifelse(cooksd2>4/sample_size, names(cooksd2),""), col="red")  # add labels

# Defining outliers based on 4/n criteria; 

influential2 <- as.numeric(names(cooksd2)[(cooksd2 > (4/sample_size2))])

# We visualize the rows that are considered outliers and will be later excluded

view(pain_data_c[influential2, ])

# We define the dataset without outliers

pain_data3 <- pain_data_c[-influential2, ]
head(pain_data3)

nrow(pain_data3)

# We convert the dataset in data frame for further anaylsis

pain_no_out2 <- as.data.frame(pain_data3)

view(pain_no_out2)

# We build the same complex model, but this time with the dataframe without outliers

c_modelupdate_no <- lm(pain ~ age + sex + STAI_trait + pain_cat + cortisol_saliva + mindfulness, data = pain_no_out2)
nrow(model.frame(c_modelupdate_no))
summary(c_modelupdate_no)


```

#### Checking assumptions of the updated model

Normality assumption

```{r}
#The Shapiro-Wilk statistic associated with these data is W .94, indicating that no significant departures from normality were detected (p 7.11).
shapiro.test(pain_no_out2$pain)

# However when plotting normal distribution seems slightly violated

qqnorm(pain_no_out2$pain,main="QQ plot of normal data",pch=19)

hist(pain_no_out2$pain,probability=T, main="Histogram of normal
data",xlab="Approximately normally distributed data")
lines(density(pain_no_out2$pain),col=2)

# Plot for detecting violation of normality assumption
ols_plot_resid_qq(c_modelupdate_no)

plot(c_modelupdate_no, 2)
# Test for detecting violation of normality assumption.

ols_test_normality(c_modelupdate_no)

```

Linearity assumption

```{r}
# Plot to check linearity

ols_plot_resid_fit(c_modelupdate_no)

plot(c_modelupdate_no, 1)

```

Homoscedasticty assumption (homogeneity of variance)

```{r}
ols_plot_resid_fit(c_modelupdate_no)

plot(c_modelupdate_no, 3)

ols_test_breusch_pagan(c_modelupdate_no)
```

Multicollinearity assumption

```{r}
ols_coll_diag(c_modelupdate_no)

ols_vif_tol(c_modelupdate_no)

vif(c_modelupdate_no)
```

## Model comparison

Create the simple model and get the results of the model that needs to be reported based on the What to report section.

The hierarchical multiple regression revealed that at Stage one, in the simple model, Age contributed
significantly to the regression model, and with Sex accounted for 6.2% of the variation in Pain
F(2, 146 = 5.92 p. 0.0034)

```{r}
s_model2 <- lm(pain ~ age + sex, data = pain_no_out2)
nrow(model.frame(s_model2))
summary(s_model2)


```

Create the more complex model based on the results of the model diagnostics. Also, get the results that needs to be reported based on the What to report section.

The complex model explained 41 % of variance. (F 6, 142 = 18.52 p.< 7.86)

```{r}
c_modelupdate_no <- lm(pain ~ age + sex + STAI_trait + pain_cat + cortisol_saliva + mindfulness, data = pain_no_out2)
nrow(model.frame(c_modelupdate_no))
summary(c_modelupdate_no)
```

Compare the two models.

```{r}
#  Nested model comparision

anova(s_model2, c_modelupdate_no)

# When comparing the two models the following results are obtained. F (2, 142 = 23.034, p. <.0001). The complex model is significantly different compared to the simple one

# We can also calculate the change in R squared when comparing the two models

summary(c_modelupdate_no)$adj.r.squared - summary(s_model2)$adj.r.squared

# Incremental variance explained in 0.35. This can be interpreted that the complex model is better at explaining the variance compared to the simple one.

pain_numeric2 <- subset(pain_no_out2, select = c(pain, STAI_trait, pain_cat, cortisol_serum, cortisol_saliva, mindfulness))

view(pain_numeric2)

pairs.panels(pain_numeric2, col = "red")


#

lrtest(s_model2, c_modelupdate_no)

#Chi squared 74.511 p < 2.528e15

AIC(s_model2, c_modelupdate_no)


```

