---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
require(data.table)
require(curl)
require(viridis)

#store the url in a variable
urlfile<-'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv'
#read it in as dataframe with curl
expeditions_data<-as.data.frame(fread(urlfile))

#factorize, filter, reorder the data and then make the plot accordingly
expeditions_data %>%
  mutate(peak_name = as.factor(peak_name)) %>%
  mutate(peak_name = forcats::fct_lump(peak_name, n = 15)) %>%
  filter(peak_name != "Other") %>%
  mutate(peak_name = fct_rev(fct_infreq(peak_name))) %>%
  ggplot( aes(x = peak_name, fill = season)) +
    geom_bar() +
    coord_flip() +
    labs(y = "Number of expeditions", title = "The 15 most popular peaks stacked by season of expedition") +
    theme_light() +
    theme(legend.position = "bottom") +
    scale_fill_viridis(discrete = TRUE)

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
#store the url in a variable
urlfile<-'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv'
#read it in as dataframe with curl
phd_data<-as.data.frame(fread(urlfile))

#summarise data and then make the plot accordingly
phd_data %>%
  mutate(broad_field = as.factor(broad_field)) %>%
  group_by(year, broad_field) %>%
  summarise(sumphd = sum(n_phds, na.rm = TRUE)) %>%
  ggplot( aes(x = year, y = sumphd, color = broad_field)) +
    geom_line(size = 1.2) +
    labs(x = element_blank(), y = element_blank(), title = "Number of awarded Ph.D.-s in the US by year", colour = "Broad field") +
    scale_fill_brewer(palette = "Dark2") +
    theme_minimal() +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    scale_y_continuous(labels = scales::comma_format())

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
#store the url in a variable
urlfile<-'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv'
#read it in as dataframe with curl
commute_data<-as.data.frame(fread(urlfile))

#summarise data, spread long to wide, make the logtransformed plot accordignly
commute_data %>%
  group_by(state, mode, state_region, state_abb) %>%
  summarise(sum_ppl = sum(n, na.rm = TRUE)) %>%
  spread(mode, sum_ppl) %>%
  ggplot( aes(x = Walk, y = Bike, color = state_region, label = state_abb)) +
    geom_point(size = 2) +
    geom_text(aes(label=state_abb), colour = "black") +
    labs(x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)", title = "Number of people walking vs. biking to work in each USA state", colour = "State region") +
    scale_x_log10(labels = scales::comma_format()) +
    scale_y_log10(labels = scales::comma_format()) +
    theme_light()
  

```
