---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(scales)
library(viridis)
library(RColorBrewer)



```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
climbing <- tt_load("2020-09-22")

expeditions_d <- climbing$expeditions

view(expeditions_d)

expeditions_d %>%
  mutate(peak_name = as.factor(peak_name)) %>%
  mutate(peak_name = forcats::fct_lump(peak_name, n = 15)) %>%
  filter(peak_name != "Other") %>%
  mutate(peak_name = fct_rev(fct_infreq(peak_name))) %>%
  ggplot( aes(y = peak_name, fill = season)) +
    geom_bar() +
    labs(x = "Number of expeditions", y = NULL, title = "The 15 most popular peaks stacked by season of expedition") +
    theme_light() +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
    scale_fill_viridis(discrete = TRUE)


```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phds <- tt_load("2019-02-19")

phds_field <- phds$phd_by_field

view(phds_field)

phds_field %>%
  group_by(broad_field, year) %>%
  summarize(sphds = sum(n_phds, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = sphds, color = broad_field)) +
  geom_line(size = 1.2) +
  labs(x = NULL, y = NULL, color = "Broad field", title = "Number of awarded Ph.D.-s in the US by year") +
  scale_x_continuous(breaks = scales::pretty_breaks()) + 
  scale_y_continuous(labels = comma_format()) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2")

view(phds_field)
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
comm_us <- tt_load("2019-11-05")

commute_us <- comm_us$commute

view(commute_us)

commute_usa <- commute_us %>%
  mutate(state = recode_factor(state, "Ca" = "California", "Massachusett" = "Massachusetts")) %>%
  group_by(state, mode, state_abb, state_region) %>%
  summarize(commutes = sum(n, na.rm = TRUE))  %>%
  spread(mode, commutes) %>%
  arrange(state_abb)
View(commute_usa)



############################
#commute_usa$state_abb<-
 # rownames(commute_usa)
#newlab<- 
 # subset(commute_usa, state_abb
 #        %in% c("AK", "AL"))

#commute_usa$Walk<-as.numeric(commute_usa$Walk)
#commute_usa$Bike<-as.numeric(commute_usa$Bike)

#ggplot(commute_usa, aes(x = Walk, y = Bike, 
 #     color = state_region)) +
#  geom_point(size = 2) +
 # scale_x_log10(labels = scales::comma_format()) +
#  scale_y_log10(labels = scales::comma_format(), limits = c(100, 200000)) +
 #    geom_text(data= newlab, color = "black", size #= 4, vjust=1.5) +
  
  #############################
  
  
  
   ggplot(commute_usa, aes(x = Walk, y = Bike, color = state_region)) +
  geom_point(size = 2) +
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format(), limits = c(100, 200000)) +
     geom_text(aes(label= state_abb), color = "black", size = 4, check_overlap = TRUE) +
 
 labs(x = "Number of ppl walking to work (log N)",
       y = "Number of ppl biking to work (log N)",
       title = "Number of people walking vs. biking to work in each USA state", color = "State region") +
  theme(aspect.ratio = 1,
       axis.title.x = element_text(size = 12.5),
       axis.title.y = element_text(size = 12.5),
        plot.title = element_text(size = 14, hjust = 0.5))  +
  theme_light()

  print(commute_usa)
  
  
#data=subset(commute_usa, Walk > 4000 | Bike > 5000)

 # mutate(state = replace(state,state == "CA", "California"))
  #subset(commute_us, mode == "Bike", select = state, state_abb, state_region)
  
  

view(commute_us)
  
  
  


```
